Stage-Splitting a Modal Language

### 2014-09-02 paper/splitting.tex
Minor changes on partial evaluation.
Lots of improvements to splitting.

Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
Despite this similarity in purpose, the form of stage-splitting is rather different. 
Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
and $f_2$ uses that data structure and the stage-\bbtwo\ input to produce the final output.  
More precisely, a stage splitter is any $s$ such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
Importantly, the stage splitting operation does not depend on the stage-\bbone\ input, unlike partial evaluation.

As with the dynamics, our stage-splitting algorithm for expressions in \lang\ take 
the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$. 
Before diving into the full generality, we consider splitting for a simple fragment of the full language: partial values.
After that, we define splitting for full expressions, discuss some notable properties of the algorithm, and then prove correctness.

@@

The basic goal of partial value splitting is to take a partial value $v$ and create from it two single stage values $v_1$ and $v_2$,
where $v_1$ contains only the information from $v$ that is relevant to the first stage, 
and $v_2$ contains only the information from $v$ that is relevant to the second stage.
There are many reasonable ways to do this, 
and we simply chose one that preserves as much structure as possible from $v$.
Formally, partial value splitting comprises a single judgment, $v \vsplito [v_1,v_2]$, the rules for which are given in \cref{fig:valSplit}.
It essentially acts distributively at product, sum, and recursive types, 
doing little more than recuring into subvalues and duplicating the structure to both results.
At base types, partial value splitting assign all content ({\em i.e.} the integer or boolean) to stage \bbone, and uses a nullary tuple placeholder for stage \bbtwo.
Conversely at $\fut$ types, splitting assigns all content ({\em i.e.} the environment reference) to stage \bbtwo, and uses a nullary tuple for stage \bbone.

@@

\subsection{Full Expression Splitting}

We consider splitting of stage \bbtwo\ terms first, as it has a simpler form.  
For example, consider \verb|hold{1+2} < 5|.
We can split this into two separate terms, \verb|1+2| and \verb|fn l => l < 5|, 
which are called the {\em precomputation} and {\em residual}\footnote{We also used the term {\em residual} for the result of partial evaluation.
Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual}
and {\em evaluation-residual}.}, respectively.

The form of splitting for stage \bbone\ terms is more complicated.
To see why, consider the term \verb|(next{hold{1+2} < 5},3*4)|.
As before, the comparison operation is part of stage \bbtwo, 
and the addition operation is a stage \bbone\ precomputation, the result of which will eventually become input to the residual.
The multiplication operation, however, is neither part of the residual nor a precomputation to support it.
Instead it is the {\em immediate result} of stage \bbone, 
because it is available to the stage \bbone\ context around our original term.
For example:
\begin{lstlisting}
let x = (next{hold{1+2} < 5},3*4) in
next{ if prev{#1 x} then 7 else hold {#2 x} }
\end{lstlisting}
To support this, stage \bbone\ splitting produces two outputs: the {\em combined result} and the residual,
where the combined result reduces to a tuple containing the immediate result and the precomputation.

We must also develop a notion of what it means for splitting to be correct.  
Our general approach is to say that the dynamic semantics from \ref{sec:semantics} and splitting method are equivalent in some way.  
In general, if a multi-stage program $P$ splits into stage-\bbone\ part $P_1$ and stage-\bbtwo\ part $P_2$, then
\begin{enumerate}
\item Partial evaluation of $P$ will terminate if and only if evaluation of $P_1$ terminates, and if so then
\item the immediate results of each method will be equivalent, and
\item the evaluation-residual and the splitting-residual ($P_2$ bound under the precomputation) are equivalent.
\end{enumerate}

This is straight-forward, given a careful definition of equivalence.

@@

\subsection {Speculation}

@@

\subsection {Stage \bbone\ Divergence}
Consider the splitting rules for stage-\bbone\ {\tt if} and {\tt case} expressions.
In both cases we have enough information at stage \bbone\ to know what branch to take, so there's no need to speculate.
Instead, we evaluate the stage \bbone\ portion of only the active branch, and then inject precomputation into a sum type.
Then in the residual, we case on that sum and resume the stage \bbtwo\ portion of the correct branch.

\subsection {Stage \bbone\ Functions}

@@

\subsection{Metatheory}

We start with two mutually dependent definitions of equivalence.  
Both relate evaluation-residuals on the left with splitting residuals on the right,
but $\equiv$ operates on terms, whereas $\sim$ on values.

@@

The jury is still out on how to make these strong enough to prove that the partitioning between stages is correct.

>>>>

Much like partial evaluation, the goal of stage-splitting is to separate the evaluation of a multi-stage function into distinct phases.
However, where partial evaluation requires the value of some inputs to be known, stage splitting is a static transformation performed
before any arguments are available.
Taking a multivariate function $f$ as input, stage-splitting produces two functions $f_1$ and $f_2$,
where $f_1$ uses the stage-\bbone\ input to produce a data structure, 
and $f_2$ uses that data structure and the stage-\bbtwo\ input to produce the final output.  
More precisely, a stage splitter is any $s$ such that
\[
	\forall f. \exists f_1,f_2. 
	\left[
		\begin{array}{l}
		s(f) = (f_1,f_2) \text{ and } \\
		\forall x,y.\llbracket f \rrbracket(x,y)=\llbracket f_2 \rrbracket(\llbracket f_1 \rrbracket(x),y)
		\end{array}
	\right]
\]
with the same $\llbracket \cdot \rrbracket$ notation as before.
Notice that $x$ here is bound underneath the existential $f_1$ and $f_2$,
meaning that the stage splitting operation does not depend on the stage-\bbone\ input.
This stands in contrast to the definition of partial evaluation, where $f_x$ does depend on $x$.

As with the dynamics, our stage-splitting algorithm for expressions in \lang\ takes 
the form of two mutually-dependent judgments, $\splitonesym$ and $\splittwosym$. 
Before diving into the full generality, we consider splitting for a simple fragment of the full language: partial values.
After that, we define splitting for full expressions, discuss some notable properties of the algorithm, and then prove correctness.

@@

The basic goal of partial value splitting is to take a partial value $v$ and create from it two single stage values $v_1$ and $v_2$,
where $v_1$ contains only the information from $v$ that is relevant to the first stage, 
and $v_2$ contains only the information from $v$ that is relevant to the second stage.
We call this action {\em masking}.

There are many reasonable ways to achieve this goal, 
and we choose one that preserves as much structure as possible from $v$.
Formally, partial value splitting comprises a single judgment, $v \vsplito [v_1,v_2]$, the rules for which are given in \cref{fig:valSplit}.
It essentially acts distributively at product, sum, and recursive types, 
doing little more than recuring into subvalues and duplicating the structure to both results.
At base types, partial value splitting assign all content ({\em i.e.} the integer or boolean) to stage \bbone, and uses a nullary tuple placeholder for stage \bbtwo.
Conversely at $\fut$ types, splitting assigns all content ({\em i.e.} the environment reference) to stage \bbtwo, and uses a nullary tuple for stage \bbone.

@@

\subsection{ Expression Splitting}

We now consider the problem of splitting expressions in \lang, starting with stage \bbtwo\ terms.  
The goal here is to end up with two new terms: one that has all of the stage \bbone\ content of the original, and one with all of the stage \bbtwo\ content. 
These are respectively called the {\em precomputation} and {\em residual}.
\footnote{We also used the term {\em residual} for the result of partial evaluation.
Where necessary, we will distinguish between these concepts by using the terms {\em splitting-residual} and {\em evaluation-residual}.}

For example, consider $e=``\verb|hold{1+2} < hold{3+4}-5|"$.
We can split this into two terms, $p=``\verb|(1+2,3+4)|"$ and $r=``\verb|fn (L1,L2) => L1 < L2-5|"$, 
with the as yet imprecise property that $r$ applied to $p$ yields the same final value as $e$.
In general, all we're doing here is finding the contained nuggets of stage \bbone\ code and pulling them into a big tuple,
an action we call {\em lifting}.

Note that lifting and masking are different, despite both being within the purview of splitting.
They particularly differ in how the resulting stage \bbone\ parts are used.
The $v_1$ of partial value splitting is essentially the stage \bbone\ content of {\em result} of computation, 
and it's accessible to whatever surrounding piece of code refers to the value being split.
In contrast, the $p$ of stage \bbtwo\ expression splitting is only meant to support the stage \bbtwo\ part,
and is not accessible externally.

The distinction between lifting and masking is important because stage \bbone\ expression splitting does both.
To see why, consider the term \verb|(next{hold{1+2} < 5},3*4)|.
The stage \bbone\ result of masking is \verb|((),3*4)|, 
and the precomputation (i.e. the stage \bbone\ result of lifting) is \verb|(1+2)|.
To get the total stage \bbone\ result of splitting, we just take the tuple of those two parts: \verb|(((),3*4),1+2)|.  
This is called the {\em combined result} for stage \bbone.
\footnote{In this example, the combined result is a 2-tuple, but more generally it only must be a term which reduces to a 2-tuple.}

The residual is likewise the combination of the stage \bbtwo\ results of lifting and masking.
In this case, it is ``\verb|fn L => (L<5,())|".

@@

\subsection {Properties}

We'll now call out a few properties to notice in the splitting rules.

\subsubsection {Speculation}

@@

\subsubsection {Stage \bbone\ Divergence}
Consider the splitting rules for stage-\bbone\ {\tt if} and {\tt case} expressions.
In both cases we have enough information at stage \bbone\ to know what branch to take, so there's no need to speculate.
Instead, we evaluate the stage \bbone\ portion of only the active branch, and then inject precomputation into a sum type.
Then in the residual, we case on that sum and resume the stage \bbtwo\ portion of the correct branch.

\subsubsection {Stage \bbone\ Functions}

@@

\subsection{Correctness of Splitting}

We can now develop an notion of what it means for splitting to be correct.  
Our general approach is to say that the dynamic semantics from \ref{sec:semantics} and splitting method are equivalent in some way.  
Thus, we start with two mutually dependent definitions of equivalence.  
Both relate evaluation-residuals on the left with splitting residuals on the right,
but $\equiv$ equates closed expressions, whereas $\sim$ equates values.

@@

### 2014-09-02 paper/techreport.tex
Lots of improvements to splitting.

With these goals in mind, our initial concern is how to express multistage programs. To this end, we have slightly modified prior work to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types. An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.

@@

The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way.  This has previously been explored as an optimization technique for languages with fewer features or a more constrained staging model than \lang. Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.

>>>>

With these goals in mind, our initial concern is how to express multistage programs. To this end, we have modified prior work to create \lang, a typed multi-stage lambda calculus featuring products, sums, and isorecursive types. An important property of \lang\ is that the staging of a well-typed program is apparent without resorting to any variable analysis, contra standard binding time analysis.

@@

The main idea of the latter technique is to statically compile the multi-stage program into several smaller programs, one per stage, and then to execute those in a standard way.  This has previously been explored as an optimization technique for languages with fewer features or a more constrained staging model than \lang. Prior examples include the ``data specialization" work of Knoblock and Ruff\,\cite{knoblock96} as well as the compilation of high-level shading languages for computer graphics such as RTSL\,\cite{Proudfoot:2001} and Spark\,\cite{Foley:2011}.  Unlike partial evaluation, stage splitting is purely a static operation and is not dependent on the input data.
